Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-01-18 paper/ms.tex
adding new matrix figure
new matrix figure in ms
incorrect factors of 2 fixed
adding transit example

\citet{Ambikasaran:2015} demonstrated two key facts that allow us to use this
extended matrix formalism in practice.
First, even if the covariance function is a mixture of exponentials, the
extended matrix will still be banded with bandwidth \dfmtodo{check}.
Second, \citet{Ambikasaran:2015} proved that the determinant of
$K_\mathrm{ext}$ is equal to the determinant of $K$ up to a sign.

>>>>

In this section, we generalize the method from the previous section to the
covariance function given by \eq{celerite-kernel}.
This derivation follows \citet{Ambikasaran:2015} but it includes explicit
treatment of complex parameters.

In the case of the full \celerite\ covariance function
(\eqalt{celerite-kernel}), we introduce the following auxiliary variables in
analogy to the $u_n$ and $g_n$ that we introduced in the previous section
\begin{eqnarray}\eqlabel{full-system-first}
\phi_{n,j} &=& e^{-c_j\,\tau_{n+1,n}}\,\cos\left(d_j\,\tau_{n+1,n}\right)\\
\psi_{n,j} &=& -e^{-c_j\,\tau_{n+1,n}}\,\sin\left(d_j\,\tau_{n+1,n}\right)\\
g_{n,j} &=& \phi_{n,j}\,g_{n-1,j} + \phi_{n,j}\,x_n + \psi_{n,j}\,h_{n-1,j}\\
h_{n,j} &=& \phi_{n,j}\,h_{n-1,j} - \psi_{n,j}\,x_n - \psi_{n,j}\,g_{n-1,j}\\
u_{n,j} &=& \phi_{n+1,j}\,u_{n+1,j} + a_j\,x_{n+1} + \psi_{n+1,j}\,v_{n+1,j}\\
v_{n,j} &=& \phi_{n+1,j}\,v_{n+1,j} - b_j\,x_{n+1} - \psi_{n+1,j}\,u_{n+1,j}
\end{eqnarray}
with the boundary conditions
\begin{eqnarray}
    g_{0,j} = 0 \quad, \quad
    h_{0,j} = 0 \quad, \quad
    u_{N,j} = 0 \quad, \quad\mathrm{and}\quad
    v_{N,j} = 0
\end{eqnarray}
for all $j$.
Using these variables and some algebra, we find that the following expression
\begin{eqnarray}\eqlabel{full-system-last}
\sum_{j=1}^J \left[a_j\,g_{n,j}+b_j\,h_{n,j}\right]
+ \left[{\sigma_n}^2+\sum_{j=1}^J a_j\right]
+ \sum_{j=1}^J \left[\phi_{n,j}\,u_{n,j}+\psi_{n,j}\,v_{n,j}\right]
    &=& y_n
\end{eqnarray}
is equivalent to the target matrix equation
\begin{eqnarray}
K\,\bvec{x} &=& \bvec{y} \quad.
\end{eqnarray}
\eq{full-system-first} through \eq{full-system-last} define a banded matrix
equation in the ``extended'' space and, as before, this can be used to solve
for $K^{-1}\,\bvec{y}$ and $\det K$ in $\mathcal{O}(J^2\,N)$ operations.
\Figure{matrix} shows a pictorial representation of the sparsity pattern of
the extended matrix $K_\mathrm{ext}$.
Given that definition of $K_\mathrm{ext}$, the corresponding extended vectors
$\bvec{x}_\mathrm{ext}$ and $\bvec{y}_\mathrm{ext}$ are defined schematically
as
\begin{eqnarray}
{\bvec{x}_\mathrm{ext}} ^\T =
\left(\begin{array}{cccccccccc}
    x_1 & u_{1,j} & v_{1,j} & g_{1,j} & h_{1,j} & x_2 & u_{2,j} & \cdots &
    h_{N-1,j} & x_N
\end{array}\right)
\end{eqnarray}
and
\begin{eqnarray}
{\bvec{y}_\mathrm{ext}} ^\T =
\left(\begin{array}{cccccccccc}
    y_1 & 0 & 0 & 0 & 0 & y_2 & 0 & \cdots &
    0 & y_N
\end{array}\right) \quad.
\end{eqnarray}

After constructing the extended matrix (using a compact storage format), the
extended matrix can be factorized using a LU-decomposition\footnote{Even
though $K_\mathrm{ext}$ is symmetric, it is not positive definite so a
Cholesky solver cannot be used for increased efficiency.} routine optimized
for band matrices.
This decomposition can be used to compute the determinant of $K$ and to solve
$K^{-1}\,\bvec{y}$.

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.7\textwidth]{figures/matrix.pdf}
\caption{A pictorial representation of the sparse extended matrix
$K_\mathrm{ext}$ with $N=5$ and $J=2$.
Each colored block corresponds to a non-zero entry in the matrix as described
in the legend.
    \figurelabel{matrix}}
\end{center}
\end{figure}
